---
title: "Lab 07 Part 2"
author: "Matthieu Huy"
date: "2023-02-23"
output: html_document
---

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(sf)
library(tmap)
library(gstat)
library(stars)
```


## Read in the California depth to groundwater data

**Data source:** California Department of Water Resources. (2016). Groundwater Depth, California, Fall 2013. California Department of Water Resources. Available at: http://purl.stanford.edu/py212tb3061.

These data are measurements for depth to groundwater recorded in 2013. We'll read in the layer with `read_sf()`, check the CRS, then make a quick plot to start

```{r}
### Read in the CA county data (TIGER shapefile):

ca_counties_sf <- read_sf(here("data/ca_counties"), 
                          layer = "CA_Counties_TIGER2016") %>% 
  janitor::clean_names() %>% 
  select(name)

### Check the projection
st_crs(ca_counties_sf) #WGS 84, epsgL3857

### Read in the depth to groundwater layer: 
ca_dgw_sf <- read_sf(here("data/ca_dgw/F2013_DBGS_Points_20150720_093252.gpkg")) %>% 
  janitor::clean_names()

### Check the projection:
st_crs(ca_dgw_sf) # WGS 84, epsg:4326

### update the groundwater data to match the counties CRS
ca_dgw_sf <- st_transform(ca_dgw_sf, st_crs(ca_counties_sf))

### can also tell it the EPSG code directly if you know it
# ca_dgw_sf <- st_transform(ca_dgw_sf, 3857)

### Make a quick ggplot:
ggplot() +
  geom_sf(data = ca_counties_sf) +
  geom_sf(data = ca_dgw_sf, aes(color = dgbs), size = 0.1) +
  scale_color_gradient(low = "pink", high = "darkred")
```

## Map these interactively to explore further

```{r}
tmap_mode("view")
tm_shape(ca_dgw_sf) +
  tm_dots("dgbs", palette = 'Blues')
### note - no CA map here, Leaflet does that for us
### check out all the palettes:
# tmaptools::palette_explorer() ### need package shinyjs
```

Let's just look at depth to groundwater (ft) in and around San Joaquin County. 

```{r}
### Just get sj_county
sj_county_sf <- ca_counties_sf %>% 
  filter(name == "San Joaquin")

### Now keep observations for groundwater depth within that county:
sj_depth_sf <- ca_dgw_sf %>% 
  st_intersection(sj_county_sf) #keep all spatial information about wells.

### Make a quick plot of those: 
ggplot() +
  geom_sf(data = sj_county_sf) +
  geom_sf(data = sj_depth_sf, aes(color = dgbs)) +
  theme_void() +
  scale_color_viridis_c()

tm_shape(sj_depth_sf) +
  tm_dots("dgbs", palette = 'Blues')
```


### Check for duplicate locations:

Interpolation chokes when observations at identical location - spatial singularity!

```{r}
well_duplicates <- sj_depth_sf %>% 
  janitor::get_dupes(latitude, longitude) #look for duplicates in latitude and longitude columns
well_duplicates

### This will cause problems (0 spatial variance). We're just going to get rid of these problematic wells recorded at the same location: 
sj_depth_sf <- sj_depth_sf %>% 
  filter(!local_well %in% well_duplicates$local_well) 
# exclude local_well value that is the same as local_well value in well_duplicates df

### Check again for duplicates: 
sj_depth_sf %>% 
  janitor::get_dupes(latitude, longitude) 
### No more duplicates
```


## Make a variogram: 

```{r}
# Create the variogram
sj_dgw_vgm <- gstat::variogram(dgbs ~ 1,
                               data = sj_depth_sf)
### Plot it
plot(sj_dgw_vgm) 

### Looks variogramish! Increasing variance as observations get further
### apart. Let's model this...

### Now create the variogram model, & see ?vgm for other model options: 
sj_dgw_vgm_fit <- fit.variogram(sj_dgw_vgm, 
                                model = vgm(model = "Gau")) ### default starting
#uses mathematical calculation to estimate variogram plot

### Plot them together:
plot(sj_dgw_vgm, sj_dgw_vgm_fit) 
### not a great fit

### Manually estimate nugget, sill, range
sj_dgw_vgm_fit <- fit.variogram(sj_dgw_vgm, 
                                model = vgm(model = "Gau", ### (leave with defaults first)
                                            nugget = 20, sill = 3000, range = 30000))
### Plot them together:
plot(sj_dgw_vgm, sj_dgw_vgm_fit) 
### You can check the sum of squares of residuals for each: 
attr(sj_dgw_vgm_fit, 'SSErr') # ...and you could compare to other models...
```


## Spatial kriging

```{r}
### Make a grid over which we'll krige: 
sj_grid <- st_bbox(sj_county_sf) %>% # create box around area of SJ county
  stars::st_as_stars(dx = 1000, dy = 1000) %>% #create a square grid within the box with pixels 1000m x 1000m
  st_crop(sj_county_sf) #crop to area of county
# sj_grid
### Look at the grid we made: 
plot(sj_grid)
### Perform ordinary kriging: 
sj_dgw_krige <- krige(dgbs ~ 1, 
                      sj_depth_sf, 
                      sj_grid, 
                      model = sj_dgw_vgm_fit)
### Initial plot of kriging outcome: 
plot(sj_dgw_krige)
### Convert it to a spatial data frame
krige_df <- as.data.frame(sj_dgw_krige) %>% 
  st_as_sf(coords = c("x","y")) %>% 
  drop_na(var1.pred)
st_crs(krige_df) <- 3857
### Then we can use ggplot: 
ggplot(data = krige_df) +
  geom_sf(aes(color = var1.pred)) +
  scale_color_gradient(low = "blue", high = "yellow")
```

...& there are ways to convert this to raster data / aggregate if you want...

## End variogram & kriging demo