---
title: 'ESM 244: Lab 07'
author: "Matthieu Huy"
date: "2023-02-23"
output: html_document
---

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(here)
library(broom)

# Spatial data packages
library(sf)
library(tmap)
```

# Part 1: Spatial Data wrangling, visualization, and a variogram

#### Today, we'll use vector data (polygons, points) to practice reading in spatial data, checking & updating the CRS, and doing some wrangling and visualization. 


We'll use several datasets:

- California County shapefile (ca_counties_tiger_2016.shp) from the US Census Bureau's 2016 MAF/TIGER database (accessed [here](https://data.ca.gov/dataset/ca-geographic-boundaries/resource/b0007416-a325-4777-9295-368ea6b710e6?inner_span=True))
- Red sesbania records (invasive species) from the CA DFW (accessed [here](https://map.dfg.ca.gov/metadata/ds0080.html))

### Read California county outlines

```{r}
ca_counties_sf <- read_sf(here('data/ca_counties/CA_Counties_TIGER2016.shp'))

ca_subset_sf <- ca_counties_sf %>% 
  janitor::clean_names() %>%
  select(county_name = name, land_area = aland)
head(ca_subset_sf) ### WARN AGAINST View()
```

What if we wanted just the dataframe, without the geometry?  Convert to dataframe and select out the geometry column:

```{r}
ca_counties_df <- ca_counties_sf %>%
  as.data.frame() %>%
  select(-geometry)
```

#### Check and set the CRS

Use `st_crs()` to check the existing CRS for spatial data. We see that this CRS is "pseudo-mercator" based on WGS 84 - primarily used for web mapping, not analysis.  WGS84 (epsg:3857), also note proj4 string and WKT definitions.

```{r}
ca_subset_sf %>% st_crs()
ggplot() +
  geom_sf(data = ca_subset_sf,
          aes(fill = land_area), #color/fill of polygons
          color = "white", #boundaires/outline of polygons
          size = 0.1) + #size of boundary lines
  theme_void() + #gets rid of all axis information
  scale_fill_gradientn(colors = c("cyan","blue","purple")) #gradientn, allows to give a vecotr of colors including middle colors
  
#notice: don't need to specify geometry column because geom_sf knows to look for these
```

### B. Invasive red sesbania records (spatial points)

Red sesbania (*Sesbania punicea*) is an invasive plant (see more information from the [California Invasive Plants Council](https://www.cal-ipc.org/plants/profile/sesbania-punicea-profile/)). Observations for locations of invasive red sesbania are from CA DFW. See metadata and information here: https://map.dfg.ca.gov/metadata/ds0080.html

The data exist in `data/red_sesbania`, and the shapefile is stored as `ds80.shp`. Let's read in the data: 

```{r}
sesbania_sf <- read_sf(here("data","red_sesbania","ds80.shp")) %>%
  janitor::clean_names()
# Check the CRS:
sesbania_sf %>% st_crs()
```

Notice that this CRS is different from the California counties CRS, so we'll want to update it to match. Use `st_transform()` to update the CRS:


```{r}
### if you know the EPSG code:
sesbania_3857_sf <- st_transform(sesbania_sf, 3857)
### if you don't know the EPSG code:
sesbania_3857_2_sf <- st_transform(sesbania_sf, st_crs(ca_counties_sf))
# Then check it: 
sesbania_3857_sf %>% st_crs()
```

#### Plot them together! 

Note: this may take a minute.  Remember, later geoms go on top.

```{r}
ggplot() +
  geom_sf(data = ca_subset_sf,
          color = "white", #boundaires/outline of polygons
          size = 0.1) +
  geom_sf(data = sesbania_3857_sf, 
          size = 1, 
          color = "red",
          alpha = 0.6) +
  theme_void()
```

#### A bit of wrangling! 

Let's say we want to find the count of red sesbania observed locations in this dataset *by county*. How can I go about joining these data so that I can find counts? Don't worry...`st_join()` has you covered for spatial joins! 

```{r}
ca_sesb_sf <- ca_subset_sf %>% 
  st_join(sesbania_3857_sf) #uses spatial info to perform join, don't need to specify mutual column
head(ca_sesb_sf)
```

And then we can find counts (note: these are not counts for individual plants, but by record in the dataset) by county.  We can't just count the rows (e.g., using count()) because some rows are counties with no records (and sesbania information is all NAs)

```{r}
sesb_counts_sf <- ca_sesb_sf %>% 
  group_by(county_name) %>%
  summarize(n_records = sum(!is.na(id))) #sum excluding NA's in id column
head(sesb_counts_sf)
```

Then we can plot a choropleth - map with outlines and shading reflecting value - using the number of records for red sesbania as the fill color (instead of what we used previously, land area):
```{r}
ggplot(data = sesb_counts_sf %>% 
            mutate(n_records = ifelse(n_records == 0, NA, n_records))) + 
  #make it so that values of 0 arent included in gradient
  geom_sf(
          aes(fill = n_records), 
          color = "white", 
          size = 0.1) +
  scale_fill_gradientn(colors = c("yellow","orange","red"), na.value = "lightgrey") +
  theme_minimal() +
  labs(fill = "Number of S. punicea records")
```

So we see that we can still use our usual wrangling skills! Let's do a bit more for fun, just to prove that our existing wrangling skills still work with spatial data - the spatial information just sticks to it! Only plot the county with the greatest number of red sesbania records (Solano), and make a map of those locations (yeah there are many ways to do this): 

```{r}
solano_sesb_sf <- ca_sesb_sf %>% 
  filter(county_name %in% "Solano")

solano_sf <- ca_subset_sf %>% 
  filter(county_name %in% "Solano")

ggplot() +
  geom_sf(data = solano_sf) +
  geom_sf(data = solano_sesb_sf,
          color = "red")
  
```
For shiny app:
reactive dropdown menu, select state or county, generate a new map for just that state or county
### C. Making an interactive map with {tmap}

Sometimes we'll want to make a map interactive so that audience members can zoom in, explore different areas, etc. We can use the {tmap} package to create an interactive map. Let's make one for our California counties (fill aesthetic by land area) with the red sesbania locations on top:

```{r}
# Set the viewing mode to "interactive":
tmap_mode(mode = "view")

# Then make a map (with the polygon fill color updated by variable 'land_area', updating the color palette to "BuGn"), then add another shape layer for the sesbania records (added as dots):
tm_shape(ca_subset_sf) +
  tm_fill("land_area", palette = "BuGn") +
  tm_shape(sesbania_sf) +
  tm_dots()
```
See *all* kinds of other cool ways you can update your interactive tmaps. 

See: 

- [`tmap` vignettes](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)
- [Chapter 8 in Robin Lovelace's "Geocomputation in R"](https://geocompr.robinlovelace.net/adv-map.html#interactive-maps) 





For Shiny App:

Reactive drop down menu where you select state/county

reactive function generates interactive tmap for just that county.

Resources for spatial data:
- RNatural Earth data: license free, no citation needed
- U.S. Census cartographic data


